---
title: "Small crustacean occupancy models"
author: "Diana Bowler, Miki Bálint"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
  html_document:
    theme: united
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: 3
---
```{r include=FALSE}
library(tidyverse)
library(plyr)
library(ggplot2)
library(reshape2)
library(unmarked)
library(boot)
```

# Crustacean OTU abundances and metadata
```{r}
rm(list = ls())
crustacean_otu_meta <- read.csv(file="crustacean_otu_meta.csv",
                                header = T, row.names = 1)
crustacean_final_replicates <- read.csv(file = "crustacean_final_replicates.csv",
                                        header = T, row.names = 1)
```

## Prepare data
```{r}
#get data files
mydata<-crustacean_final_replicates
mydata <- data.frame(X = rownames(crustacean_final_replicates),
                     crustacean_final_replicates)

#melt the data frame. very long, load the data instead
mydata<-melt(mydata,id="X")
names(mydata)<-c("OTU","Sample","Count")

#separate the sample columns
mydata$Rep<-sapply(as.character(mydata$Sample),function(x)strsplit(x,"_")[[1]][2])
mydata$Lake<-sapply(as.character(mydata$Sample),function(x)strsplit(x,"\\.")[[1]][2])
mydata$Depth<-sapply(as.character(mydata$Sample),function(x)strsplit(x,"\\.")[[1]][3])
mydata$Depth<-as.numeric(sapply(mydata$Depth,function(x)strsplit(x,"_")[[1]][1]))
```

## Remove OTU seen less than 7x
```{r}
#summary statistics of the OTUs
sumStats<-ddply(mydata,.(OTU),summarise,
                nu=length(Count[Count>0&!is.na(Count)]))
head(sumStats)
mydata<-subset(mydata,!mydata$OTU%in%sumStats$OTU[sumStats$nu < 7])
```

# Plotting the data
```{r fig.asp=1.3}
ggplot(data=mydata,
       aes(x=Depth,y=Count))+
  scale_y_log10()+
  geom_point(aes(colour=Rep))+
  facet_grid(OTU~Lake)
```

# Fit site-occupancy model 

## Format data for unmarked

For simplicity combine data across lakes
```{r}
mydata_combined <-
  ddply(mydata,
        .(OTU,Depth,Rep),
        summarise,
        Count=sum(Count))
```

Using only presence/absence data
```{r}

mydata_combined$Count <- 
  ifelse(mydata_combined$Count>0,1,0)

```

## Fit combined lake model species by species and estimate detection and occupancy at each depth

```{r}
mydataAll<-ddply(mydata_combined,
                 .(OTU),
                 function(x){
  
  #reformatting for the package
  mydataD<-dcast(x,Depth~Rep,value.var="Count")
  y = mydataD[,c("A","B","C","D","E","F")]
  siteCovs = data.frame(Depth=mydataD[,"Depth"])
  obsCovs =  list(depth=mydataD[,c("Depth","Depth","Depth","Depth","Depth","Depth")])
  
  #define the dataframe for the package
  wt <- unmarkedFrameOccu(y = y, siteCovs = siteCovs, obsCovs =  obsCovs)
  
  #using pcount
  ## Two models: Detection covariates follow first tilde, then Osccupancy covariates are second
  fm1 <- occu(~depth ~Depth, data=wt)
  
  #take the fitted relationship of detectability with with depth
  my_intercept<-as.numeric(fm1@estimates@estimates$det@estimates[1])
  my_slope<-as.numeric(fm1@estimates@estimates$det@estimates[2])
  x$predictedDetection<-inv.logit(my_intercept+my_slope*x$Depth)
  
  #get predicted occupancy with depth
  my_intercept<-as.numeric(fm1@estimates@estimates$state@estimates[1])
  my_slope<-as.numeric(fm1@estimates@estimates$state@estimates[2])
  x$predictedOccupancy<-inv.logit(my_intercept+my_slope*x$Depth)
  
  return(x)
})
```

### Plot depth-dependent detection and occupancy

Relationship between depth and detection
```{r}
qplot(Depth,predictedDetection,data=mydataAll,colour=OTU)+
  theme(legend.position="none")
```

Relationship between depth and occupancy
```{r}
qplot(Depth,predictedOccupancy,data=mydataAll,colour=OTU)+
  theme(legend.position="none")
```

## Fit single lake model species by species and estimate detection and occupancy at each depth

Diana, I don't know how to pack everything into a single function... probably would make sense.

Miki, ok!

Using only presence/absence data:

```{r}
mydata$Count <- 
  ifelse(mydata$Count>0,1,0)
```

Check first that all OTUs are in multiple Lakes (we cant fit a Lake effect if not!!)

```{r}
sumStats<-ddply(mydata,.(OTU),summarise,nuLakes=length(unique(Lake[Count!=0])))
head(sumStats)
```

Just to get something working - let just use OTUs seen in at least 6 lakes (later will can relax this and
even drop the "lake effect" for OTU seen in only one lake)

```{r}

mydata<-subset(mydata,OTU%in%sumStats$OTU[sumStats$nuLakes>5])
length(unique(mydata$OTU))

# just 5??
# yes, most of the OTUs were quite rare.
# my explanation - the eDNA detects up small crustaceans only if biomasses are high enough, and that seems to happen when the trophic conditions of the lake are high.

```

Now fit the model to this small dataset..

```{r}
  #reformatting for the package

lakeModels<-dlply(mydata,.(OTU),function(x){

mydataD_lake<-dcast(x,
                      Depth + Lake ~ Rep,
                      value.var="Count")

  y = mydataD_lake[,c("A","B","C","D","E","F")]
  siteCovs = data.frame(Depth = mydataD_lake[,"Depth"], 
                        Lake = mydataD_lake[,"Lake"])
  obsCovs =  list(depth=mydataD_lake[,c("Depth","Depth",
                                      "Depth","Depth",
                                      "Depth","Depth")])
  
  #define the dataframe for the package
  wt <- unmarkedFrameOccu(y = y, 
                          siteCovs = siteCovs, 
                          obsCovs =  obsCovs)
  
  #using pcount
  ## Two models: Detection covariates follow first tilde, then Osccupancy covariates are second
  fm1_lake <- occu(~depth ~Depth + Lake, data=wt)
  
  #lets look at the summary - here you can see all the things that can be extracted
  summary(fm1_lake)
  
  #take the fitted relationship of detectability with with depth
  my_intercept_lake <- 
    as.numeric(fm1_lake@estimates@estimates$det@estimates[1])
  my_slope_lake <- 
    as.numeric(fm1_lake@estimates@estimates$det@estimates[2])
  x$predictedDetection <-
    inv.logit(my_intercept_lake+my_slope_lake*x$Depth)
  
  #get predicted occupancy with depth
  my_intercept_lake <-
    as.numeric(fm1_lake@estimates@estimates$state@estimates[1])
  my_slope_lake<-as.numeric(fm1_lake@estimates@estimates$state@estimates[2])
  x$predictedOccupancy <-
    inv.logit(my_intercept_lake+my_slope_lake*x$Depth)
  
  #get predicted effects of lake (differences between reference lake and the others)
  #identify reference lake
  refLake<-sort(unique(x$Lake))[1]
  nuLakes<-length(unique(x$Lake))#so we know how many coefficients to extract
  myEstimates=fm1_lake@estimates@estimates$state@estimates[3:(nuLakes+1)]
  #package up the results as a data frame
  lakeEffects<-data.frame(OTU=unique(x$OTU),
                          refLake = refLake,
                          LakeDiff=names(myEstimates),
                          Coef=as.numeric(myEstimates))
  
  #we have two outputs
  
  return(list(x,lakeEffects))
  
  })
  
```

The above gave us a list with 2 sets of results per OTU

Pull out the depth results first

```{r}
mydata<-ldply(lakeModels,function(x){
        x[[1]]
})

```


### Plot each lake depth and detection

Depth and detection
```{r}

qplot(Depth,predictedDetection,data=mydata,colour=OTU)+
  theme(legend.position="none")+facet_wrap(~Lake)

```

Relationship between depth and occupancy
```{r}
qplot(Depth,predictedOccupancy,data=mydata,colour=OTU)+
  theme(legend.position="none") + facet_wrap(~Lake)
```

Pull out the lake results - you can decide how to plot it!!

```{r}
lakeOutput<-ldply(lakeModels,function(x){
        x[[2]]
})

head(lakeOutput)
```

